---
title: "04_2 Test7 actual mock benchmark"
author: "Test7 decontamination analysis"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: bootstrap
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 11, fig.height = 6.5, dpi = 180
)
options(stringsAsFactors = FALSE)
```

# 1. Scope

This benchmark uses the observed Test7 mock samples directly. The three Cell
Mock samples and the DNA Mock are not simulated. An observed ASV assigned to
one of the six MSA-2005 target genera is treated as expected; any other observed
ASV is treated as off-target. Thus, the labels are operational taxonomic truth,
not proof of the physical contamination source.

The contaminant calls are the pooled results already generated in script 06.
Mocks were excluded when those methods were fitted, and the resulting calls are
applied here to the mock samples as an external validation set.

```{r packages-paths}
library(phyloseq)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(pROC)

input_path <- file.path("Phyloseq", "phy_F270R210_260921.rds")
method_dir <- file.path(
  "Results", "Test7", "06_combined_cross_method_validation"
)
output_dir <- file.path(
  "Results", "Test7", "04_2_actual_mock_benchmark"
)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

decontam_flag_path <- file.path(method_dir, "06_05_decontam_feature_flags.csv")
microbiem_flag_path <- file.path(
  method_dir, "06_15_MicrobIEM_threshold_feature_flags.csv"
)
scrub_flag_path <- file.path(
  method_dir, "06_17_SCRuB_nonspatial_threshold_feature_flags.csv"
)

stopifnot(
  file.exists(input_path),
  file.exists(decontam_flag_path),
  file.exists(microbiem_flag_path),
  file.exists(scrub_flag_path)
)

method_colors <- c(
  Prevalence = "#1B9E77",
  Frequency = "#D95F02",
  Combined = "#7570B3",
  MicrobIEM = "#66A61E",
  `SCRuB non-spatial` = "#E7298A"
)
theme_set(theme_bw(base_size = 11))
```

# 2. Actual mock samples and ASV truth

```{r actual-mock-truth}
expected_genera <- c(
  "Acinetobacter", "Corynebacterium", "Micrococcus",
  "Cutibacterium", "Staphylococcus", "Streptococcus"
)

expected_table <- data.frame(
  Genus = expected_genera,
  expected_abundance = rep(1 / length(expected_genera), length(expected_genera))
)

phy <- readRDS(input_path)
metadata <- data.frame(sample_data(phy), check.names = FALSE) |>
  tibble::rownames_to_column("SampleID") |>
  mutate(Cycle = sub("\\.0$", "", as.character(Cycle)))

cell_mock_ids <- metadata |>
  filter(!is.na(control_type), control_type == "Cell_mix", Cycle == "3") |>
  pull(SampleID)
dna_mock_ids <- metadata |>
  filter(!is.na(control_type), control_type == "DNA_mix") |>
  pull(SampleID)
mock_ids <- c(cell_mock_ids, dna_mock_ids)

count_table <- as(otu_table(phy), "matrix")
if (taxa_are_rows(phy)) count_table <- t(count_table)
count_table <- count_table[sample_names(phy), taxa_names(phy), drop = FALSE]

taxonomy <- data.frame(tax_table(phy), check.names = FALSE) |>
  tibble::rownames_to_column("ASV") |>
  mutate(
    Genus = if_else(is.na(Genus) | Genus == "", "Unclassified", as.character(Genus))
  )

mock_truth <- as.data.frame(as.table(count_table[mock_ids, , drop = FALSE])) |>
  rename(SampleID = Var1, ASV = Var2, reads = Freq) |>
  mutate(SampleID = as.character(SampleID), ASV = as.character(ASV)) |>
  filter(reads > 0) |>
  left_join(
    metadata |> select(SampleID, Subject, control_type),
    by = "SampleID"
  ) |>
  left_join(taxonomy, by = "ASV") |>
  mutate(
    mock_type = if_else(control_type == "Cell_mix", "Cell Mock", "DNA Mock"),
    truth = if_else(Genus %in% expected_genera, "Expected", "Off-target"),
    is_off_target = truth == "Off-target"
  )

mock_truth_summary <- mock_truth |>
  group_by(mock_type, SampleID, Subject) |>
  summarise(
    total_reads = sum(reads),
    observed_ASVs = n(),
    expected_ASVs = sum(!is_off_target),
    off_target_ASVs = sum(is_off_target),
    off_target_read_pct = 100 * sum(reads[is_off_target]) / sum(reads),
    .groups = "drop"
  )

write_csv(expected_table, file.path(output_dir, "04_2_01_expected_genera.csv"))
write_csv(mock_truth, file.path(output_dir, "04_2_02_actual_mock_ASV_truth.csv"))
write_csv(mock_truth_summary, file.path(output_dir, "04_2_03_actual_mock_summary.csv"))

knitr::kable(mock_truth_summary, digits = 3)
```

# 3. Existing five-method calls

No method is refitted in this document. It uses the feature-level calls from
script 06 so that all methods use the same Test7 input and negative controls.
For MicrobIEM, the no-well-position setting (`Any detected`) is used. For
SCRuB, the non-spatial result is used.

```{r assemble-predictions}
mock_asvs <- unique(mock_truth$ASV)

decontam_flags <- read_csv(decontam_flag_path, show_col_types = FALSE) |>
  filter(batch_mode == "pooled") |>
  transmute(
    method = recode(
      method,
      prevalence = "Prevalence",
      frequency = "Frequency",
      combined = "Combined"
    ),
    threshold_parameter = "decontam threshold",
    threshold = as.numeric(threshold),
    ASV = feature_id,
    predicted_contaminant = contaminant
  )

decontam_settings <- decontam_flags |>
  distinct(method, threshold_parameter, threshold)

decontam_predictions <- tidyr::crossing(
  decontam_settings,
  ASV = mock_asvs
) |>
  left_join(
    decontam_flags,
    by = c("method", "threshold_parameter", "threshold", "ASV")
  ) |>
  mutate(predicted_contaminant = replace_na(predicted_contaminant, FALSE))

microbiem_flags <- read_csv(microbiem_flag_path, show_col_types = FALSE) |>
  filter(span_label == "Any detected") |>
  transmute(
    method = "MicrobIEM",
    threshold_parameter = "ratio threshold",
    threshold = as.numeric(ratio_threshold),
    ASV,
    predicted_contaminant = contaminant
  )

microbiem_settings <- microbiem_flags |>
  distinct(method, threshold_parameter, threshold)

microbiem_predictions <- tidyr::crossing(
  microbiem_settings,
  ASV = mock_asvs
) |>
  left_join(
    microbiem_flags,
    by = c("method", "threshold_parameter", "threshold", "ASV")
  ) |>
  mutate(predicted_contaminant = replace_na(predicted_contaminant, FALSE))

scrub_flags <- read_csv(scrub_flag_path, show_col_types = FALSE) |>
  transmute(
    method = "SCRuB non-spatial",
    threshold_parameter = "ASV removal-fraction cutoff",
    threshold = as.numeric(removal_fraction_cutoff),
    ASV,
    predicted_contaminant = affected
  )

scrub_settings <- scrub_flags |>
  distinct(method, threshold_parameter, threshold)

scrub_predictions <- tidyr::crossing(
  scrub_settings,
  ASV = mock_asvs
) |>
  left_join(
    scrub_flags,
    by = c("method", "threshold_parameter", "threshold", "ASV")
  ) |>
  mutate(predicted_contaminant = replace_na(predicted_contaminant, FALSE))

predictions <- bind_rows(
  decontam_predictions,
  microbiem_predictions,
  scrub_predictions
)
```

# 4. Classification metrics on actual mocks

The evaluation unit is an observed ASV-sample pair. Consequently, an ASV
observed in all three Cell Mock samples contributes three observations, whereas
an ASV observed only once contributes one.

```{r classification-metrics}
prediction_units <- predictions |>
  inner_join(
    mock_truth |>
      select(SampleID, Subject, mock_type, ASV, reads, Genus, truth, is_off_target),
    by = "ASV"
  )

threshold_metrics <- prediction_units |>
  group_by(mock_type, method, threshold_parameter, threshold) |>
  summarise(
    TP = sum(predicted_contaminant & is_off_target),
    TN = sum(!predicted_contaminant & !is_off_target),
    FP = sum(predicted_contaminant & !is_off_target),
    FN = sum(!predicted_contaminant & is_off_target),
    off_target_reads_removed_pct = 100 *
      sum(reads[predicted_contaminant & is_off_target]) /
      sum(reads[is_off_target]),
    expected_reads_removed_pct = 100 *
      sum(reads[predicted_contaminant & !is_off_target]) /
      sum(reads[!is_off_target]),
    .groups = "drop"
  ) |>
  mutate(
    Sensitivity_Recall = TP / (TP + FN),
    Specificity = TN / (TN + FP),
    Precision_PPV = if_else(TP + FP > 0, TP / (TP + FP), NA_real_),
    F1 = if_else(
      is.finite(Precision_PPV) & Precision_PPV + Sensitivity_Recall > 0,
      2 * Precision_PPV * Sensitivity_Recall /
        (Precision_PPV + Sensitivity_Recall),
      0
    ),
    Youden_J = Sensitivity_Recall + Specificity - 1
  )

raw_metrics <- mock_truth |>
  group_by(mock_type) |>
  summarise(
    method = "Raw: no ASV removed",
    TP = 0L,
    TN = sum(!is_off_target),
    FP = 0L,
    FN = sum(is_off_target),
    Sensitivity_Recall = 0,
    Specificity = 1,
    Precision_PPV = NA_real_,
    F1 = 0,
    Youden_J = 0,
    .groups = "drop"
  )

write_csv(threshold_metrics, file.path(output_dir, "04_2_04_threshold_metrics.csv"))
write_csv(raw_metrics, file.path(output_dir, "04_2_05_raw_classification_metrics.csv"))
```

# 5. Threshold-free AUROC

Continuous scores are reconstructed from each method's saved threshold
trajectory. ASVs absent from a method's fitted feature table receive score 0
because the fitted method cannot flag them.

```{r auc}
decontam_scores <- decontam_predictions |>
  group_by(method, ASV) |>
  summarise(
    score = if_else(
      any(predicted_contaminant),
      1 - min(threshold[predicted_contaminant]),
      0
    ),
    .groups = "drop"
  )

microbiem_scores <- microbiem_predictions |>
  group_by(method, ASV) |>
  summarise(
    score = if_else(
      any(predicted_contaminant),
      max(threshold[predicted_contaminant]),
      0
    ),
    .groups = "drop"
  )

scrub_scores <- scrub_predictions |>
  group_by(method, ASV) |>
  summarise(
    score = if_else(
      any(predicted_contaminant),
      max(threshold[predicted_contaminant]),
      0
    ),
    .groups = "drop"
  )

score_units <- bind_rows(decontam_scores, microbiem_scores, scrub_scores) |>
  inner_join(
    mock_truth |> select(SampleID, mock_type, ASV, is_off_target),
    by = "ASV"
  )

auc_parts <- list()
roc_parts <- list()
part_index <- 0L

for (mock_name in c("Cell Mock", "DNA Mock")) {
  for (method_name in names(method_colors)) {
    score_data <- score_units |>
      filter(mock_type == mock_name, method == method_name)

    roc_fit <- pROC::roc(
      response = score_data$is_off_target,
      predictor = score_data$score,
      levels = c(FALSE, TRUE),
      direction = "<",
      quiet = TRUE
    )

    part_index <- part_index + 1L
    auc_parts[[part_index]] <- data.frame(
      mock_type = mock_name,
      method = method_name,
      AUROC = as.numeric(pROC::auc(roc_fit))
    )

    roc_coordinates <- data.frame(pROC::coords(
      roc_fit,
      x = "all",
      ret = c("specificity", "sensitivity"),
      transpose = FALSE
    )) |>
      mutate(
        mock_type = mock_name,
        method = method_name,
        false_positive_rate = 1 - specificity
      )
    roc_parts[[part_index]] <- roc_coordinates
  }
}

auc_summary <- bind_rows(auc_parts)
roc_data <- bind_rows(roc_parts)

best_metrics <- threshold_metrics |>
  group_by(mock_type, method) |>
  arrange(desc(Youden_J), desc(F1), threshold, .by_group = TRUE) |>
  slice_head(n = 1) |>
  ungroup() |>
  left_join(auc_summary, by = c("mock_type", "method")) |>
  mutate(
    mock_type = factor(mock_type, levels = c("Cell Mock", "DNA Mock")),
    method = factor(method, levels = names(method_colors))
  ) |>
  arrange(mock_type, method)

write_csv(auc_summary, file.path(output_dir, "04_2_06_AUROC.csv"))
write_csv(best_metrics, file.path(output_dir, "04_2_07_best_Youden_summary.csv"))
write_csv(roc_data, file.path(output_dir, "04_2_08_ROC_coordinates.csv"))

knitr::kable(
  best_metrics |>
    select(
      mock_type, method, threshold_parameter, threshold,
      TP, TN, FP, FN, Sensitivity_Recall, Specificity,
      Precision_PPV, F1, Youden_J, AUROC,
      off_target_reads_removed_pct, expected_reads_removed_pct
    ),
  digits = 3
)
```

# 6. Figures

```{r metric-plot, fig.width=13, fig.height=9}
metric_plot_data <- best_metrics |>
  select(
    mock_type, method,
    Sensitivity_Recall, Specificity, Precision_PPV, F1, Youden_J, AUROC
  ) |>
  pivot_longer(
    cols = c(
      Sensitivity_Recall, Specificity, Precision_PPV, F1, Youden_J, AUROC
    ),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(
    metric = recode(
      metric,
      Sensitivity_Recall = "Sensitivity / Recall",
      Precision_PPV = "Precision / PPV",
      Youden_J = "Youden J"
    ),
    metric = factor(
      metric,
      levels = c(
        "Sensitivity / Recall", "Specificity", "Precision / PPV",
        "F1", "Youden J", "AUROC"
      )
    )
  )

p_metrics <- ggplot(
  metric_plot_data,
  aes(method, value, color = method)
) +
  geom_point(size = 2.7) +
  facet_grid(metric ~ mock_type) +
  scale_color_manual(values = method_colors) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Five-method performance on actual Test7 mocks",
    subtitle = "Threshold-dependent metrics use maximum Youden J; AUROC is threshold-free",
    x = NULL, y = "Metric value", color = NULL
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    legend.position = "bottom"
  )

print(p_metrics)
ggsave(
  file.path(output_dir, "04_2_01_actual_mock_metrics.png"),
  p_metrics, width = 13, height = 9, dpi = 300
)
```

```{r roc-plot, fig.width=12, fig.height=5.5}
roc_plot_data <- roc_data |>
  mutate(
    mock_type = factor(mock_type, levels = c("Cell Mock", "DNA Mock")),
    method = factor(method, levels = names(method_colors))
  )

p_roc <- ggplot(
  roc_plot_data,
  aes(false_positive_rate, sensitivity, color = method, linetype = method)
) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey65") +
  geom_path(linewidth = 0.9) +
  facet_wrap(~mock_type, nrow = 1) +
  scale_color_manual(values = method_colors) +
  coord_equal() +
  labs(
    title = "Actual-mock ASV classification ROC curves",
    x = "False-positive rate (1 - specificity)",
    y = "Sensitivity",
    color = NULL,
    linetype = NULL
  ) +
  theme(legend.position = "bottom")

print(p_roc)
ggsave(
  file.path(output_dir, "04_2_02_actual_mock_ROC.png"),
  p_roc, width = 12, height = 5.5, dpi = 300
)
```

# 7. Verification and interpretation boundary

```{r verification}
mock_unit_counts <- mock_truth |>
  count(mock_type, name = "expected_units")

metric_unit_check <- threshold_metrics |>
  mutate(observed_units = TP + TN + FP + FN) |>
  left_join(mock_unit_counts, by = "mock_type")

verification <- data.frame(
  check = c(
    "Three Cell Mock samples and one DNA Mock were included",
    "Both expected and off-target ASVs occur in each mock type",
    "All five methods produced threshold results",
    "All five methods produced AUROC for both mock types",
    "Every confusion matrix contains every observed mock ASV-sample unit",
    "All metrics are within their valid ranges"
  ),
  passed = c(
    length(cell_mock_ids) == 3L & length(dna_mock_ids) == 1L,
    all((mock_truth |> count(mock_type, truth) |> count(mock_type))$n == 2L),
    setequal(unique(threshold_metrics$method), names(method_colors)),
    nrow(auc_summary) == 10L & all(is.finite(auc_summary$AUROC)),
    all(metric_unit_check$observed_units == metric_unit_check$expected_units),
    all(is.na(threshold_metrics$Sensitivity_Recall) |
          threshold_metrics$Sensitivity_Recall >= 0 &
          threshold_metrics$Sensitivity_Recall <= 1) &
      all(is.na(threshold_metrics$Specificity) |
            threshold_metrics$Specificity >= 0 &
            threshold_metrics$Specificity <= 1) &
      all(is.na(threshold_metrics$Precision_PPV) |
            threshold_metrics$Precision_PPV >= 0 &
            threshold_metrics$Precision_PPV <= 1) &
      all(is.na(threshold_metrics$F1) |
            threshold_metrics$F1 >= 0 & threshold_metrics$F1 <= 1) &
      all(threshold_metrics$Youden_J >= -1 & threshold_metrics$Youden_J <= 1) &
      all(auc_summary$AUROC >= 0 & auc_summary$AUROC <= 1)
  )
)

write_csv(verification, file.path(output_dir, "04_2_09_verification.csv"))
knitr::kable(verification)
if (!all(verification$passed)) stop("One or more actual-mock checks failed.")
```

Sensitivity here means removal of observed off-target ASVs. Specificity means
retention of observed ASVs assigned to the six expected genera. A flagged ASV
within an expected genus is counted as a false positive even if that particular
ASV is not the intended strain, which is a conservative genus-level definition.
The DNA Mock contains one sample, so its uncertainty must not be interpreted as
equivalent to replicated validation.

```{r session-info}
sessionInfo()
```
