---
title: "04. Mock community validation"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
    df_print: paged
params:
  force_recompute: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 300
)
```

```{r packages}
# 분석에 사용하는 패키지와 공통 함수를 불러온다.
source("./load_packages.R")
source("./functions.R")

package_dir <- normalizePath("../../decontamSensitivity", winslash = "/")
devtools::load_all(package_dir, quiet = TRUE)

input_path <- "./Phyloseq/phy_F270R210_260921.rds"
results_dir <- "./Results"
dir.create(results_dir, showWarnings = FALSE)

thresholds <- c(0.01, 0.05, seq(0.1, 0.9, by = 0.1))
methods <- c("prevalence", "frequency")
control_scenarios <- c(
  "prep_control_only",
  "air_swab_only",
  "prep_and_air_swab"
)

scenario_label <- c(
  prep_control_only = "Prep control only",
  air_swab_only = "Air Swab only",
  prep_and_air_swab = "Prep control + Air Swab"
)

method_label <- c(
  prevalence = "Prevalence",
  frequency = "Frequency"
)
```

# 1) Mock community and reference profiles

## MSA-2005 theoretical composition

```{r theoretical-composition}
# 제공된 MSA-2005 균주를 Genus 수준의 균등 조성으로 정의한다.
mock_truth <- tibble::tribble(
  ~Genus, ~Species, ~ATCC, ~expected_abundance,
  "Acinetobacter", "Acinetobacter johnsonii", "ATCC 17909", 1 / 6,
  "Corynebacterium", "Corynebacterium striatum", "ATCC 6940", 1 / 6,
  "Micrococcus", "Micrococcus luteus", "ATCC 4698", 1 / 6,
  "Cutibacterium", "Cutibacterium acnes", "ATCC 11828", 1 / 6,
  "Staphylococcus", "Staphylococcus epidermidis", "ATCC 12228", 1 / 6,
  "Streptococcus", "Streptococcus mitis", "ATCC 49456", 1 / 6
)

readr::write_csv(
  mock_truth,
  file.path(results_dir, "04_MSA2005_theoretical_composition.csv")
)

knitr::kable(mock_truth, digits = 4)
```

## Sample assignment

```{r sample-assignment}
phy <- readRDS(input_path)

metadata <- data.frame(sample_data(phy), check.names = FALSE) |>
  tibble::rownames_to_column("SampleID") |>
  dplyr::mutate(
    Cycle = sub("\\.0$", "", as.character(Cycle)),
    is_prep_control = !is.na(control_type) & control_type == "Prep control",
    is_air_swab = !is.na(control_type) & control_type == "Air Swab",
    is_cell_mock = !is.na(control_type) & control_type == "Cell_mix",
    is_dna_mock = !is.na(control_type) & control_type == "DNA_mix",
    is_positive_control = !is.na(Sample_Control) & Sample_Control == "PC",
    is_true = !is_prep_control & !is_air_swab & !is_positive_control,
    concentration_raw = as.numeric(DNA_ng_uL)
  )

zero_replacement <- min(
  metadata$concentration_raw[metadata$concentration_raw > 0],
  na.rm = TRUE
) / 2

metadata <- metadata |>
  dplyr::mutate(
    decontam_concentration = dplyr::if_else(
      concentration_raw == 0,
      zero_replacement,
      concentration_raw
    )
  )

phy_metadata <- metadata |>
  tibble::column_to_rownames("SampleID")
sample_data(phy) <- sample_data(phy_metadata)

cell_mock_ids <- metadata |>
  dplyr::filter(is_cell_mock, Cycle == "3") |>
  dplyr::pull(SampleID)

dna_mock_id <- metadata |>
  dplyr::filter(is_dna_mock) |>
  dplyr::pull(SampleID)

mock_sample_table <- metadata |>
  dplyr::filter(is_cell_mock | is_dna_mock) |>
  dplyr::select(
    SampleID,
    Subject,
    Cycle,
    control_type,
    DNA_ng_uL,
    Mol_Conc_nM,
    Total_uL
  )

readr::write_csv(
  mock_sample_table,
  file.path(results_dir, "04_mock_sample_assignment.csv")
)

knitr::kable(mock_sample_table, digits = 3)
```

The Cell mock has three Cycle 3 replicates. The DNA mock was sequenced in the same run and is used as an observed reference. Because two target organisms are not consistently resolved to Species in the taxonomy table, accuracy is evaluated at Genus level.

## Raw mock profiles

```{r raw-mock-profile}
count_table <- as(otu_table(phy), "matrix")
if (taxa_are_rows(phy)) {
  count_table <- t(count_table)
}

taxonomy <- data.frame(tax_table(phy), check.names = FALSE) |>
  tibble::rownames_to_column("feature_id") |>
  dplyr::mutate(
    Genus = dplyr::if_else(
      is.na(Genus) | Genus == "",
      "Unclassified",
      as.character(Genus)
    )
  )

genus_by_asv <- taxonomy$Genus[match(colnames(count_table), taxonomy$feature_id)]
names(genus_by_asv) <- colnames(count_table)
```

```{r profile-function}
# Raw와 decontam 결과에서 반복하여 사용하는 Genus profile을 계산한다.
make_genus_profile <- function(sample_id, removed_asv = character()) {
  sample_count <- count_table[sample_id, ]
  sample_count[names(sample_count) %in% removed_asv] <- 0

  genus_count <- rowsum(
    matrix(sample_count, ncol = 1),
    group = genus_by_asv[names(sample_count)],
    reorder = FALSE
  )[, 1]

  tibble::tibble(
    SampleID = sample_id,
    Genus = names(genus_count),
    count = as.numeric(genus_count),
    relative_abundance = if (sum(genus_count) > 0) {
      as.numeric(genus_count) / sum(genus_count)
    } else {
      rep(0, length(genus_count))
    }
  )
}
```

```{r raw-reference-profiles}
raw_cell_list <- vector("list", length(cell_mock_ids))
for (i in seq_along(cell_mock_ids)) {
  raw_cell_list[[i]] <- make_genus_profile(cell_mock_ids[i])
}
raw_cell_profile <- dplyr::bind_rows(raw_cell_list)

dna_mock_profile <- make_genus_profile(dna_mock_id) |>
  dplyr::select(Genus, dna_mock_abundance = relative_abundance)

raw_cell_mean <- raw_cell_profile |>
  dplyr::group_by(Genus) |>
  dplyr::summarise(
    relative_abundance = mean(relative_abundance),
    .groups = "drop"
  )

reference_composition <- dplyr::full_join(
  mock_truth |>
    dplyr::select(Genus, theoretical_abundance = expected_abundance),
  dna_mock_profile,
  by = "Genus"
) |>
  dplyr::full_join(
    raw_cell_mean |>
      dplyr::rename(raw_cell_abundance = relative_abundance),
    by = "Genus"
  ) |>
  dplyr::mutate(
    dplyr::across(
      c(theoretical_abundance, dna_mock_abundance, raw_cell_abundance),
      ~ tidyr::replace_na(.x, 0)
    )
  ) |>
  dplyr::arrange(dplyr::desc(dna_mock_abundance))

readr::write_csv(
  reference_composition,
  file.path(results_dir, "04_reference_composition.csv")
)

reference_composition |>
  dplyr::filter(
    theoretical_abundance > 0 |
      dna_mock_abundance >= 0.001 |
      raw_cell_abundance >= 0.001
  ) |>
  knitr::kable(digits = 4)
```

```{r reference-composition-plot, fig.width=11, fig.height=6}
target_genera <- mock_truth$Genus

reference_plot_data <- reference_composition |>
  dplyr::mutate(
    Genus_plot = dplyr::if_else(Genus %in% target_genera, Genus, "Other")
  ) |>
  dplyr::select(
    Genus_plot,
    Theoretical = theoretical_abundance,
    `DNA mock` = dna_mock_abundance,
    `Raw Cell mock` = raw_cell_abundance
  ) |>
  tidyr::pivot_longer(
    cols = -Genus_plot,
    names_to = "reference",
    values_to = "relative_abundance"
  ) |>
  dplyr::group_by(reference, Genus_plot) |>
  dplyr::summarise(
    relative_abundance = sum(relative_abundance),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    reference = factor(
      reference,
      levels = c("Theoretical", "DNA mock", "Raw Cell mock")
    )
  )

mock_taxa_color <- c(
  "Acinetobacter" = "#1B9E77",
  "Corynebacterium" = "#D95F02",
  "Micrococcus" = "#7570B3",
  "Cutibacterium" = "#E7298A",
  "Staphylococcus" = "#66A61E",
  "Streptococcus" = "#E6AB02",
  "Other" = "grey80"
)

reference_composition_plot <- ggplot(
  reference_plot_data,
  aes(x = reference, y = relative_abundance, fill = Genus_plot)
) +
  geom_col(width = 0.75) +
  scale_fill_manual(values = mock_taxa_color) +
  scale_y_continuous(labels = scales::label_percent(), expand = c(0, 0)) +
  labs(x = NULL, y = "Relative abundance", fill = "Genus") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

ggsave(
  file.path(results_dir, "04_reference_composition.png"),
  reference_composition_plot,
  width = 11,
  height = 6,
  dpi = 300
)

reference_composition_plot
```

# 2) Cycle 3 decontam analysis

```{r cycle3-balance}
# Positive control은 contaminant 판정에서 제외하고 검증에만 사용한다.
cycle3_balance <- metadata |>
  dplyr::filter(Cycle == "3") |>
  dplyr::mutate(
    sample_role = dplyr::case_when(
      is_true ~ "True",
      is_prep_control ~ "Prep control",
      is_air_swab ~ "Air Swab",
      is_cell_mock ~ "Cell mock",
      TRUE ~ "Other"
    )
  ) |>
  dplyr::count(sample_role, name = "n_samples")

knitr::kable(cycle3_balance)
```

```{r cycle3-decontam-function}
# Cycle 3에서 control 정의와 method 조합별 contaminant ASV를 판정한다.
run_cycle3_decontam <- function(control_scenario, method) {
  meta <- data.frame(sample_data(phy), check.names = FALSE)

  selected_control <- switch(
    control_scenario,
    prep_control_only = meta$is_prep_control,
    air_swab_only = meta$is_air_swab,
    prep_and_air_swab = meta$is_prep_control | meta$is_air_swab
  )

  keep_sample <- meta$Cycle == "3" & (meta$is_true | selected_control)
  run_phy <- prune_samples(keep_sample, phy)
  run_phy <- prune_taxa(taxa_sums(run_phy) > 0, run_phy)

  run_meta <- data.frame(sample_data(run_phy), check.names = FALSE)
  run_meta$decontam_group <- ifelse(
    run_meta$is_prep_control | run_meta$is_air_swab,
    "control",
    "biological"
  )
  sample_data(run_phy) <- sample_data(run_meta)

  argument <- list(
    ps = run_phy,
    control_column = "decontam_group",
    control_label = "control",
    thresholds = thresholds,
    method = method
  )

  if (method == "frequency") {
    argument$concentration_column <- "decontam_concentration"
  }

  list(
    control_scenario = control_scenario,
    method = method,
    result = do.call(
      decontamSensitivity::run_decontam_threshold_sweep,
      argument
    )
  )
}
```

```{r run-cycle3-decontam}
analysis_grid <- tidyr::crossing(
  control_scenario = control_scenarios,
  method = methods
) |>
  dplyr::mutate(
    run_id = paste(control_scenario, method, sep = "__")
  )

cache_path <- file.path(results_dir, "04_cycle3_decontam_runs.rds")
cache_config <- list(
  input_md5 = unname(tools::md5sum(input_path)),
  thresholds = thresholds,
  methods = methods,
  control_scenarios = control_scenarios
)

use_cache <- FALSE
if (file.exists(cache_path) && !isTRUE(params$force_recompute)) {
  cached <- readRDS(cache_path)
  use_cache <- identical(cached$config, cache_config)
}

if (use_cache) {
  cycle3_runs <- cached$runs
} else {
  cycle3_runs <- vector("list", nrow(analysis_grid))
  names(cycle3_runs) <- analysis_grid$run_id

  for (i in seq_len(nrow(analysis_grid))) {
    run_info <- analysis_grid[i, ]
    cycle3_runs[[run_info$run_id]] <- run_cycle3_decontam(
      control_scenario = run_info$control_scenario,
      method = run_info$method
    )
  }

  saveRDS(
    list(config = cache_config, runs = cycle3_runs),
    cache_path
  )
}

analysis_grid |>
  dplyr::mutate(cache_used = use_cache) |>
  knitr::kable()
```

```{r collect-cycle3-flags}
flag_list <- vector("list", length(cycle3_runs))

for (i in seq_along(cycle3_runs)) {
  run <- cycle3_runs[[i]]

  flag_list[[i]] <- run$result$feature_flags |>
    dplyr::filter(contaminant) |>
    dplyr::mutate(
      control_scenario = run$control_scenario,
      control_definition = unname(scenario_label[run$control_scenario]),
      method = run$method,
      method_definition = unname(method_label[run$method]),
      .before = 1
    )
}

cycle3_flags <- dplyr::bind_rows(flag_list) |>
  dplyr::arrange(control_scenario, method, threshold, feature_id)

readr::write_csv(
  cycle3_flags,
  file.path(results_dir, "04_cycle3_flagged_ASV.csv")
)
```

# 3) Cell mock profiles after decontamination

```{r filtered-cell-profiles}
profile_list <- vector(
  "list",
  nrow(analysis_grid) * length(thresholds) * length(cell_mock_ids)
)
profile_index <- 1

for (run_id in names(cycle3_runs)) {
  run <- cycle3_runs[[run_id]]

  for (threshold_value in thresholds) {
    removed_asv <- cycle3_flags |>
      dplyr::filter(
        control_scenario == run$control_scenario,
        method == run$method,
        threshold == threshold_value
      ) |>
      dplyr::pull(feature_id)

    for (sample_id in cell_mock_ids) {
      profile_list[[profile_index]] <- make_genus_profile(
        sample_id,
        removed_asv
      ) |>
        dplyr::mutate(
          control_scenario = run$control_scenario,
          control_definition = unname(scenario_label[run$control_scenario]),
          method = run$method,
          method_definition = unname(method_label[run$method]),
          threshold = round(threshold_value, 2),
          removed_ASVs = length(removed_asv),
          .before = 1
        )

      profile_index <- profile_index + 1
    }
  }
}

filtered_cell_profile <- dplyr::bind_rows(profile_list)

readr::write_csv(
  filtered_cell_profile,
  file.path(results_dir, "04_filtered_cell_mock_genus_profile.csv")
)
```

## Composition at threshold 0.5

```{r threshold-05-composition, fig.width=15, fig.height=10}
composition_05 <- filtered_cell_profile |>
  dplyr::filter(threshold == 0.5) |>
  dplyr::mutate(
    Genus_plot = dplyr::if_else(Genus %in% target_genera, Genus, "Other")
  ) |>
  dplyr::group_by(
    control_definition,
    method_definition,
    SampleID,
    Genus_plot
  ) |>
  dplyr::summarise(
    relative_abundance = sum(relative_abundance),
    .groups = "drop"
  )

composition_05_plot <- ggplot(
  composition_05,
  aes(x = SampleID, y = relative_abundance, fill = Genus_plot)
) +
  geom_col(width = 0.8) +
  facet_grid(method_definition ~ control_definition) +
  scale_fill_manual(values = mock_taxa_color) +
  scale_y_continuous(labels = scales::label_percent(), expand = c(0, 0)) +
  labs(
    x = NULL,
    y = "Relative abundance after decontamination",
    fill = "Genus"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "white", color = "grey50"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6),
    panel.spacing = grid::unit(4, "mm"),
    legend.position = "right"
  )

ggsave(
  file.path(results_dir, "04_cell_mock_composition_threshold_0.5.png"),
  composition_05_plot,
  width = 15,
  height = 10,
  dpi = 300
)

composition_05_plot
```

# 4) Accuracy metrics

The accuracy metrics follow the ITS2 mock benchmark: detection based F1 score, precision, recall and false discovery rate, together with abundance based R squared and Bray Curtis distance. Higher F1 and R squared and lower Bray Curtis distance indicate closer agreement with the reference.

```{r accuracy-function}
# 각 profile을 theoretical composition과 DNA mock에 반복 비교한다.
calculate_accuracy <- function(profile_df, expected_df, reference_name) {
  genus_universe <- union(profile_df$Genus, expected_df$Genus)

  observed <- profile_df$relative_abundance[match(genus_universe, profile_df$Genus)]
  expected <- expected_df$expected_abundance[match(genus_universe, expected_df$Genus)]
  observed[is.na(observed)] <- 0
  expected[is.na(expected)] <- 0

  detected <- observed > 0
  expected_detected <- expected > 0
  TP <- sum(detected & expected_detected)
  FP <- sum(detected & !expected_detected)
  FN <- sum(!detected & expected_detected)

  precision <- ifelse(TP + FP > 0, TP / (TP + FP), NA_real_)
  recall <- ifelse(TP + FN > 0, TP / (TP + FN), NA_real_)
  F1 <- ifelse(
    !is.na(precision) & !is.na(recall) & precision + recall > 0,
    2 * precision * recall / (precision + recall),
    NA_real_
  )

  tibble::tibble(
    reference = reference_name,
    TP = TP,
    FP = FP,
    FN = FN,
    precision = precision,
    recall = recall,
    F1 = F1,
    false_discovery_rate = ifelse(TP + FP > 0, FP / (TP + FP), NA_real_),
    R2 = ifelse(
      sd(observed) > 0 & sd(expected) > 0,
      cor(observed, expected)^2,
      NA_real_
    ),
    Bray_Curtis = as.numeric(
      vegan::vegdist(rbind(observed, expected), method = "bray")[1]
    ),
    target_abundance = sum(observed[genus_universe %in% target_genera]),
    off_target_abundance = sum(observed[!genus_universe %in% target_genera])
  )
}
```

```{r calculate-accuracy}
theoretical_reference <- mock_truth |>
  dplyr::select(Genus, expected_abundance)

dna_reference <- dna_mock_profile |>
  dplyr::rename(expected_abundance = dna_mock_abundance)

profile_setting <- filtered_cell_profile |>
  dplyr::distinct(
    control_scenario,
    control_definition,
    method,
    method_definition,
    threshold,
    SampleID
  )

accuracy_list <- vector("list", nrow(profile_setting) * 2)
accuracy_index <- 1

for (i in seq_len(nrow(profile_setting))) {
  setting <- profile_setting[i, ]

  profile_i <- filtered_cell_profile |>
    dplyr::filter(
      control_scenario == setting$control_scenario,
      method == setting$method,
      threshold == setting$threshold,
      SampleID == setting$SampleID
    ) |>
    dplyr::select(Genus, relative_abundance)

  accuracy_list[[accuracy_index]] <- calculate_accuracy(
    profile_i,
    theoretical_reference,
    "MSA-2005 theoretical"
  ) |>
    dplyr::bind_cols(setting)
  accuracy_index <- accuracy_index + 1

  accuracy_list[[accuracy_index]] <- calculate_accuracy(
    profile_i,
    dna_reference,
    "DNA mock observed"
  ) |>
    dplyr::bind_cols(setting)
  accuracy_index <- accuracy_index + 1
}

accuracy_metrics <- dplyr::bind_rows(accuracy_list) |>
  dplyr::select(
    control_scenario,
    control_definition,
    method,
    method_definition,
    threshold,
    SampleID,
    dplyr::everything()
  )

# 제거 전 Cell mock의 정확도를 baseline으로 계산한다.
raw_accuracy_list <- vector("list", length(cell_mock_ids) * 2)
raw_accuracy_index <- 1

for (sample_id in cell_mock_ids) {
  raw_profile_i <- raw_cell_profile |>
    dplyr::filter(SampleID == sample_id) |>
    dplyr::select(Genus, relative_abundance)

  raw_accuracy_list[[raw_accuracy_index]] <- calculate_accuracy(
    raw_profile_i,
    theoretical_reference,
    "MSA-2005 theoretical"
  ) |>
    dplyr::mutate(SampleID = sample_id)
  raw_accuracy_index <- raw_accuracy_index + 1

  raw_accuracy_list[[raw_accuracy_index]] <- calculate_accuracy(
    raw_profile_i,
    dna_reference,
    "DNA mock observed"
  ) |>
    dplyr::mutate(SampleID = sample_id)
  raw_accuracy_index <- raw_accuracy_index + 1
}

raw_accuracy_metrics <- dplyr::bind_rows(raw_accuracy_list)

raw_accuracy_summary <- raw_accuracy_metrics |>
  dplyr::group_by(reference) |>
  dplyr::summarise(
    dplyr::across(
      c(
        precision,
        recall,
        F1,
        false_discovery_rate,
        R2,
        Bray_Curtis,
        target_abundance,
        off_target_abundance
      ),
      ~ mean(.x, na.rm = TRUE),
      .names = "{.col}_raw"
    ),
    .groups = "drop"
  )

accuracy_summary <- accuracy_metrics |>
  dplyr::group_by(
    reference,
    control_scenario,
    control_definition,
    method,
    method_definition,
    threshold
  ) |>
  dplyr::summarise(
    dplyr::across(
      c(
        precision,
        recall,
        F1,
        false_discovery_rate,
        R2,
        Bray_Curtis,
        target_abundance,
        off_target_abundance
      ),
      list(mean = ~ mean(.x, na.rm = TRUE), se = ~ sd(.x, na.rm = TRUE) / sqrt(dplyr::n())),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"
  )

readr::write_csv(
  accuracy_metrics,
  file.path(results_dir, "04_mock_accuracy_by_sample.csv")
)
readr::write_csv(
  accuracy_summary,
  file.path(results_dir, "04_mock_accuracy_summary.csv")
)
readr::write_csv(
  raw_accuracy_metrics,
  file.path(results_dir, "04_raw_mock_accuracy_by_sample.csv")
)
readr::write_csv(
  raw_accuracy_summary,
  file.path(results_dir, "04_raw_mock_accuracy_summary.csv")
)
```

## Main accuracy metrics

```{r main-accuracy-plot, fig.width=15, fig.height=11}
main_metric_plot_data <- accuracy_summary |>
  dplyr::select(
    reference,
    control_definition,
    method_definition,
    threshold,
    F1_mean,
    R2_mean,
    Bray_Curtis_mean
  ) |>
  tidyr::pivot_longer(
    cols = c(F1_mean, R2_mean, Bray_Curtis_mean),
    names_to = "metric",
    values_to = "value"
  ) |>
  dplyr::mutate(
    metric = dplyr::recode(
      metric,
      F1_mean = "F1 score",
      R2_mean = "R squared",
      Bray_Curtis_mean = "Bray Curtis distance"
    ),
    metric = factor(
      metric,
      levels = c("F1 score", "R squared", "Bray Curtis distance")
    )
  )

raw_main_plot_data <- raw_accuracy_summary |>
  dplyr::select(reference, F1_raw, R2_raw, Bray_Curtis_raw) |>
  tidyr::pivot_longer(
    cols = c(F1_raw, R2_raw, Bray_Curtis_raw),
    names_to = "metric",
    values_to = "value"
  ) |>
  dplyr::mutate(
    metric = dplyr::recode(
      metric,
      F1_raw = "F1 score",
      R2_raw = "R squared",
      Bray_Curtis_raw = "Bray Curtis distance"
    ),
    metric = factor(
      metric,
      levels = c("F1 score", "R squared", "Bray Curtis distance")
    )
  )

main_accuracy_plot <- ggplot(
  main_metric_plot_data,
  aes(x = threshold, y = value, color = method_definition)
) +
  geom_hline(
    data = raw_main_plot_data,
    aes(yintercept = value),
    inherit.aes = FALSE,
    linetype = 2,
    color = "grey35",
    linewidth = 0.6
  ) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  facet_grid(
    reference + metric ~ control_definition,
    scales = "free_y"
  ) +
  scale_color_manual(values = c(
    "Prevalence" = "#0072B2",
    "Frequency" = "#D55E00"
  )) +
  scale_x_continuous(breaks = thresholds) +
  labs(x = "Threshold", y = "Accuracy metric", color = "Method") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "white", color = "grey50"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6),
    panel.spacing = grid::unit(4, "mm"),
    legend.position = "top"
  )

ggsave(
  file.path(results_dir, "04_mock_accuracy_main_metrics.png"),
  main_accuracy_plot,
  width = 15,
  height = 11,
  dpi = 300
)

main_accuracy_plot
```

## Detection metrics

```{r detection-metric-plot, fig.width=15, fig.height=10}
detection_plot_data <- accuracy_summary |>
  dplyr::select(
    reference,
    control_definition,
    method_definition,
    threshold,
    precision_mean,
    recall_mean,
    false_discovery_rate_mean
  ) |>
  tidyr::pivot_longer(
    cols = c(precision_mean, recall_mean, false_discovery_rate_mean),
    names_to = "metric",
    values_to = "value"
  ) |>
  dplyr::mutate(
    metric = dplyr::recode(
      metric,
      precision_mean = "Precision",
      recall_mean = "Recall",
      false_discovery_rate_mean = "False discovery rate"
    )
  )

raw_detection_plot_data <- raw_accuracy_summary |>
  dplyr::select(
    reference,
    precision_raw,
    recall_raw,
    false_discovery_rate_raw
  ) |>
  tidyr::pivot_longer(
    cols = c(precision_raw, recall_raw, false_discovery_rate_raw),
    names_to = "metric",
    values_to = "value"
  ) |>
  dplyr::mutate(
    metric = dplyr::recode(
      metric,
      precision_raw = "Precision",
      recall_raw = "Recall",
      false_discovery_rate_raw = "False discovery rate"
    )
  )

detection_metric_plot <- ggplot(
  detection_plot_data,
  aes(x = threshold, y = value, color = method_definition)
) +
  geom_hline(
    data = raw_detection_plot_data,
    aes(yintercept = value),
    inherit.aes = FALSE,
    linetype = 2,
    color = "grey35",
    linewidth = 0.6
  ) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  facet_grid(
    reference + metric ~ control_definition,
    scales = "free_y"
  ) +
  scale_color_manual(values = c(
    "Prevalence" = "#0072B2",
    "Frequency" = "#D55E00"
  )) +
  scale_x_continuous(breaks = thresholds) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Threshold", y = "Detection metric", color = "Method") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "white", color = "grey50"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6),
    panel.spacing = grid::unit(4, "mm"),
    legend.position = "top"
  )

ggsave(
  file.path(results_dir, "04_mock_accuracy_detection_metrics.png"),
  detection_metric_plot,
  width = 15,
  height = 10,
  dpi = 300
)

detection_metric_plot
```

## Best threshold by Bray Curtis distance

```{r best-threshold}
# 각 reference에서 평균 Bray Curtis distance가 가장 낮은 조건을 제시한다.
best_threshold <- accuracy_summary |>
  dplyr::group_by(reference, control_definition, method_definition) |>
  dplyr::slice_min(Bray_Curtis_mean, n = 1, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::select(
    reference,
    control_definition,
    method_definition,
    threshold,
    Bray_Curtis_mean,
    R2_mean,
    F1_mean,
    target_abundance_mean,
    off_target_abundance_mean
  ) |>
  dplyr::left_join(raw_accuracy_summary, by = "reference") |>
  dplyr::mutate(
    Bray_Curtis_improvement = Bray_Curtis_raw - Bray_Curtis_mean,
    R2_improvement = R2_mean - R2_raw,
    F1_improvement = F1_mean - F1_raw
  ) |>
  dplyr::select(
    reference,
    control_definition,
    method_definition,
    threshold,
    Bray_Curtis_mean,
    Bray_Curtis_raw,
    Bray_Curtis_improvement,
    R2_mean,
    R2_raw,
    R2_improvement,
    F1_mean,
    F1_raw,
    F1_improvement,
    target_abundance_mean,
    off_target_abundance_mean
  ) |>
  dplyr::arrange(reference, control_definition, method_definition)

readr::write_csv(
  best_threshold,
  file.path(results_dir, "04_best_threshold_by_reference.csv")
)

knitr::kable(best_threshold, digits = 4)
```

## Accuracy at threshold 0.5

```{r threshold-05-accuracy}
accuracy_threshold_05 <- accuracy_summary |>
  dplyr::filter(threshold == 0.5) |>
  dplyr::left_join(raw_accuracy_summary, by = "reference") |>
  dplyr::mutate(
    Bray_Curtis_improvement = Bray_Curtis_raw - Bray_Curtis_mean,
    R2_improvement = R2_mean - R2_raw,
    F1_improvement = F1_mean - F1_raw
  ) |>
  dplyr::select(
    reference,
    control_definition,
    method_definition,
    threshold,
    Bray_Curtis_mean,
    Bray_Curtis_raw,
    Bray_Curtis_improvement,
    R2_mean,
    R2_raw,
    R2_improvement,
    F1_mean,
    F1_raw,
    F1_improvement
  ) |>
  dplyr::arrange(reference, control_definition, method_definition)

readr::write_csv(
  accuracy_threshold_05,
  file.path(results_dir, "04_accuracy_threshold_0.5_vs_raw.csv")
)

knitr::kable(accuracy_threshold_05, digits = 4)
```

The theoretical and DNA mock references answer different questions. Agreement with the theoretical profile evaluates recovery of the nominal equal mixture, whereas agreement with the DNA mock evaluates whether Cell mock processing approaches the composition observed after bypassing cell lysis and DNA extraction. The two results should therefore be reported separately.

```{r session-info}
sessionInfo()
```
