---
title: "Karstens et al. 2019: three-level decontamination benchmark"
author: "decontamSensitivity project"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: bootstrap
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 12, fig.height = 7, dpi = 180
)
options(stringsAsFactors = FALSE)
set.seed(2019)
```

# 1. Benchmark scope

This analysis uses the processed even mock-community dilution data and the
sequence-level gold standard released by Karstens et al. (2019). The dataset
contains nine three-fold mock dilutions (`D0`--`D8`), one extraction blank,
1,414 ASVs, nine expected mock ASVs, and 1,405 contaminant ASVs.

The contingency-table convention follows the public MicrobIEM benchmarking
scripts: a contaminant read removed is a true positive (TP), an expected mock
read retained is a true negative (TN), an expected mock read removed is a false
positive (FP), and a contaminant read retained is a false negative (FN).

```{r packages-paths}
library(phyloseq)
library(decontam)
library(SCRuB)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(patchwork)

input_dir <- file.path(
  "Input", "Test11_Karstens2019", "Karstens2019_official"
)
input_path <- file.path(input_dir, "mockDilutionsPrep.RData")
microbiem_source <- paste0(
  "D:/KSY/Project/10.AT_multicenter/Results/260909_reanalysis/",
  "tmp/tool_sources/MicrobIEM/MicrobIEM_decontamination.R"
)
output_dir <- file.path("Results", "Karstens2019", "07_three_level_benchmark")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

stopifnot(file.exists(input_path), file.exists(microbiem_source))
source(microbiem_source)

method_colors <- c(
  Prevalence = "#1B9E77",
  Frequency = "#D95F02",
  Combined = "#7570B3",
  MicrobIEM = "#66A61E",
  `SCRuB non-spatial` = "#E7298A"
)
intensity_levels <- c("Default / reference", "Intermediate", "Aggressive")
theme_set(theme_bw(base_size = 11))
```

# 2. Input and gold standard

```{r load-input}
data_environment <- new.env(parent = emptyenv())
load(input_path, envir = data_environment)
ps <- data_environment$ps
mock_taxa <- as.character(data_environment$mock_taxa)

count_matrix <- as(otu_table(ps), "matrix")
if (taxa_are_rows(ps)) count_matrix <- t(count_matrix)
count_matrix <- count_matrix[sample_names(ps), taxa_names(ps), drop = FALSE]

sample_ids <- paste0("D", 0:8)
control_id <- "Blank"
asv_ids <- colnames(count_matrix)

sample_metadata <- data.frame(sample_data(ps), check.names = FALSE) |>
  tibble::rownames_to_column("SampleID") |>
  mutate(
    dilution_index = as.integer(sub("D", "", SampleID)),
    input_cells = if_else(
      SampleID == control_id,
      0,
      1.5e9 / 3^dilution_index
    ),
    dilution = factor(SampleID, levels = sample_ids)
  )

truth_table <- data.frame(
  ASV = asv_ids,
  truth = ifelse(asv_ids %in% mock_taxa, "Expected mock", "Contaminant"),
  is_contaminant = !asv_ids %in% mock_taxa
)

count_long <- as.data.frame(as.table(count_matrix[sample_ids, , drop = FALSE])) |>
  rename(SampleID = Var1, ASV = Var2, reads = Freq) |>
  mutate(SampleID = as.character(SampleID), ASV = as.character(ASV)) |>
  left_join(truth_table, by = "ASV") |>
  left_join(
    sample_metadata |> select(SampleID, DNA_conc, input_cells, dilution),
    by = "SampleID"
  )

composition_summary <- count_long |>
  group_by(SampleID, dilution, input_cells, truth) |>
  summarise(reads = sum(reads), .groups = "drop") |>
  group_by(SampleID) |>
  mutate(relative_abundance = reads / sum(reads)) |>
  ungroup()

input_summary <- composition_summary |>
  group_by(SampleID, dilution, input_cells) |>
  summarise(
    total_reads = sum(reads),
    contaminant_read_pct = 100 * sum(relative_abundance[truth == "Contaminant"]),
    .groups = "drop"
  )

write_csv(truth_table, file.path(output_dir, "07_01_gold_standard_ASVs.csv"))
write_csv(input_summary, file.path(output_dir, "07_02_input_summary.csv"))
knitr::kable(input_summary, digits = 3)
```

# 3. Three removal-intensity levels

The three levels represent increasing removal intensity, not identical numeric
parameters across tools. `decontam` uses its documented default threshold 0.1,
then 0.5 and 0.9. MicrobIEM uses published ratio thresholds 1, 0.5, and 0.1;
smaller ratios remove more features. SCRuB has no equivalent classification
threshold, so its non-spatial model is fitted once and ASVs are classified using
post-hoc sample-level removal-fraction cutoffs 0.5, 0.2, and 0.05.

```{r parameter-map}
parameter_map <- tibble::tribble(
  ~method, ~intensity, ~parameter, ~value,
  "Prevalence", "Default / reference", "decontam threshold", 0.1,
  "Prevalence", "Intermediate", "decontam threshold", 0.5,
  "Prevalence", "Aggressive", "decontam threshold", 0.9,
  "Frequency", "Default / reference", "decontam threshold", 0.1,
  "Frequency", "Intermediate", "decontam threshold", 0.5,
  "Frequency", "Aggressive", "decontam threshold", 0.9,
  "Combined", "Default / reference", "decontam threshold", 0.1,
  "Combined", "Intermediate", "decontam threshold", 0.5,
  "Combined", "Aggressive", "decontam threshold", 0.9,
  "MicrobIEM", "Default / reference", "negative-control ratio", 1,
  "MicrobIEM", "Intermediate", "negative-control ratio", 0.5,
  "MicrobIEM", "Aggressive", "negative-control ratio", 0.1,
  "SCRuB non-spatial", "Default / reference", "removal-fraction cutoff", 0.5,
  "SCRuB non-spatial", "Intermediate", "removal-fraction cutoff", 0.2,
  "SCRuB non-spatial", "Aggressive", "removal-fraction cutoff", 0.05
) |>
  mutate(
    method = factor(method, levels = names(method_colors)),
    intensity = factor(intensity, levels = intensity_levels)
  )

write_csv(parameter_map, file.path(output_dir, "07_03_parameter_map.csv"))
knitr::kable(parameter_map)
```

# 4. Run decontamination methods

All methods are fitted once to the complete dilution experiment. This matches
the intended use case of selecting one analysis setting for one dataset. Mock
samples are biological samples and `Blank` is the negative control.

```{r run-decontam}
prediction_parts <- list()
prediction_index <- 0L

for (method_code in c("prevalence", "frequency", "combined")) {
  method_label <- recode(
    method_code,
    prevalence = "Prevalence",
    frequency = "Frequency",
    combined = "Combined"
  )

  method_settings <- parameter_map |>
    filter(method == method_label)

  for (setting_index in seq_len(nrow(method_settings))) {
    setting <- method_settings[setting_index, ]
    arguments <- list(
      seqtab = ps,
      method = method_code,
      threshold = setting$value
    )
    if (method_code %in% c("prevalence", "combined")) {
      arguments$neg <- "is.neg"
    }
    if (method_code %in% c("frequency", "combined")) {
      arguments$conc <- "DNA_conc"
    }

    fit <- do.call(decontam::isContaminant, arguments)
    feature_flags <- data.frame(
      ASV = rownames(fit),
      predicted_contaminant = fit$contaminant %in% TRUE
    )

    prediction_index <- prediction_index + 1L
    prediction_parts[[prediction_index]] <- tidyr::crossing(
      SampleID = sample_ids,
      ASV = asv_ids
    ) |>
      left_join(feature_flags, by = "ASV") |>
      mutate(
        predicted_contaminant = replace_na(predicted_contaminant, FALSE),
        method = method_label,
        intensity = as.character(setting$intensity),
        parameter = as.character(setting$parameter),
        parameter_value = setting$value
      )
  }
}
```

```{r run-microbiem}
microbiem_settings <- parameter_map |>
  filter(method == "MicrobIEM")

for (setting_index in seq_len(nrow(microbiem_settings))) {
  setting <- microbiem_settings[setting_index, ]
  fit <- MicrobIEM_decontamination(
    feature_table = data.frame(t(count_matrix), check.names = FALSE),
    SAMPLE = sample_ids,
    NEG1 = control_id,
    ratio_NEG1_threshold = setting$value,
    ratio_NEG2_threshold = NA_real_,
    span_NEG1_threshold = NA_real_,
    span_NEG2_threshold = NA_real_
  )

  feature_flags <- data.frame(
    ASV = rownames(fit),
    predicted_contaminant = fit$is_contaminant %in% TRUE
  )

  prediction_index <- prediction_index + 1L
  prediction_parts[[prediction_index]] <- tidyr::crossing(
    SampleID = sample_ids,
    ASV = asv_ids
  ) |>
    left_join(feature_flags, by = "ASV") |>
    mutate(
      predicted_contaminant = replace_na(predicted_contaminant, FALSE),
      method = "MicrobIEM",
      intensity = as.character(setting$intensity),
      parameter = as.character(setting$parameter),
      parameter_value = setting$value
    )
}
```

```{r run-scrub}
scrub_metadata <- data.frame(
  is_control = rownames(count_matrix) == control_id,
  control_type = ifelse(rownames(count_matrix) == control_id, "Blank", "Sample"),
  row.names = rownames(count_matrix)
)

scrub_fit <- SCRuB::SCRuB(
  input_data = count_matrix,
  metadata = scrub_metadata,
  control_order = "Blank",
  verbose = FALSE
)

scrub_clean <- scrub_fit$decontaminated_samples[sample_ids, asv_ids, drop = FALSE]
scrub_original <- count_matrix[sample_ids, asv_ids, drop = FALSE]
scrub_removal_fraction <- matrix(
  0,
  nrow = length(sample_ids),
  ncol = length(asv_ids),
  dimnames = list(sample_ids, asv_ids)
)
present_entries <- scrub_original > 0
scrub_removal_fraction[present_entries] <- pmax(
  scrub_original[present_entries] - scrub_clean[present_entries],
  0
) / scrub_original[present_entries]

scrub_settings <- parameter_map |>
  filter(method == "SCRuB non-spatial")

for (setting_index in seq_len(nrow(scrub_settings))) {
  setting <- scrub_settings[setting_index, ]
  prediction_index <- prediction_index + 1L
  prediction_parts[[prediction_index]] <- as.data.frame(
    as.table(scrub_removal_fraction >= setting$value)
  ) |>
    rename(SampleID = Var1, ASV = Var2, predicted_contaminant = Freq) |>
    mutate(
      SampleID = as.character(SampleID),
      ASV = as.character(ASV),
      method = "SCRuB non-spatial",
      intensity = as.character(setting$intensity),
      parameter = as.character(setting$parameter),
      parameter_value = setting$value
    )
}

predictions <- bind_rows(prediction_parts) |>
  mutate(
    method = factor(method, levels = names(method_colors)),
    intensity = factor(intensity, levels = intensity_levels)
  )
```

# 5. Read-weighted benchmark metrics

```{r metrics}
benchmark_units <- predictions |>
  inner_join(
    count_long |>
      select(SampleID, ASV, reads, truth, is_contaminant, input_cells, dilution),
    by = c("SampleID", "ASV")
  )

benchmark_metrics <- benchmark_units |>
  group_by(
    SampleID, dilution, input_cells,
    method, intensity, parameter, parameter_value
  ) |>
  summarise(
    TP = sum(reads[predicted_contaminant & is_contaminant]),
    TN = sum(reads[!predicted_contaminant & !is_contaminant]),
    FP = sum(reads[predicted_contaminant & !is_contaminant]),
    FN = sum(reads[!predicted_contaminant & is_contaminant]),
    n_flagged_ASVs = sum(predicted_contaminant),
    .groups = "drop"
  ) |>
  mutate(
    Sensitivity_Recall = TP / (TP + FN),
    Specificity = TN / (TN + FP),
    Precision_PPV = if_else(TP + FP > 0, TP / (TP + FP), 0),
    F1 = if_else(
      Precision_PPV + Sensitivity_Recall > 0,
      2 * Precision_PPV * Sensitivity_Recall /
        (Precision_PPV + Sensitivity_Recall),
      0
    ),
    Youden_J = Sensitivity_Recall + Specificity - 1,
    contaminant_reads_removed_pct = 100 * Sensitivity_Recall,
    expected_reads_retained_pct = 100 * Specificity,
    total_reads_removed_pct = 100 * (TP + FP) / (TP + TN + FP + FN)
  )

published_focus <- c("D2", "D4", "D6", "D8")
focus_summary <- benchmark_metrics |>
  filter(SampleID %in% published_focus) |>
  group_by(method, intensity, parameter, parameter_value) |>
  summarise(
    mean_Sensitivity_Recall = mean(Sensitivity_Recall),
    mean_Specificity = mean(Specificity),
    mean_Precision_PPV = mean(Precision_PPV),
    mean_F1 = mean(F1),
    mean_Youden_J = mean(Youden_J),
    mean_total_reads_removed_pct = mean(total_reads_removed_pct),
    .groups = "drop"
  ) |>
  arrange(intensity, desc(mean_Youden_J))

write_csv(benchmark_metrics, file.path(output_dir, "07_04_sample_metrics.csv"))
write_csv(focus_summary, file.path(output_dir, "07_05_D2_D4_D6_D8_summary.csv"))

knitr::kable(focus_summary, digits = 3)
```

# 6. Figure 1: mock benchmark and method differences

```{r figure-1, fig.width=16, fig.height=12}
composition_plot <- composition_summary |>
  mutate(
    truth = factor(truth, levels = c("Expected mock", "Contaminant")),
    dilution = factor(SampleID, levels = sample_ids)
  ) |>
  ggplot(aes(dilution, relative_abundance, fill = truth)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = c("Expected mock" = "#0072B2", "Contaminant" = "#D55E00")) +
  scale_y_continuous(labels = scales::label_percent(), expand = c(0, 0)) +
  labs(
    x = "Three-fold dilution",
    y = "Read relative abundance",
    fill = NULL,
    title = "Observed composition across the Karstens dilution series"
  ) +
  theme(legend.position = "bottom")

youden_plot <- benchmark_metrics |>
  ggplot(aes(dilution, Youden_J, color = method, group = method)) +
  geom_hline(yintercept = 0, color = "grey60", linetype = 2) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~intensity, nrow = 1) +
  scale_color_manual(values = method_colors) +
  scale_y_continuous(limits = c(-1, 1)) +
  labs(
    x = "Three-fold dilution",
    y = "Youden J",
    color = NULL,
    title = "Classification performance changes with method and removal intensity"
  ) +
  theme(legend.position = "bottom")

tradeoff_plot <- focus_summary |>
  ggplot(aes(mean_Specificity, mean_Sensitivity_Recall, color = method)) +
  geom_point(size = 3) +
  facet_wrap(~intensity, nrow = 1) +
  scale_color_manual(values = method_colors) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Expected-read retention (specificity)",
    y = "Contaminant-read removal (sensitivity)",
    color = NULL,
    title = "Removal–retention trade-off in D2, D4, D6, and D8"
  ) +
  theme(legend.position = "bottom")

figure_1 <- composition_plot /
  (youden_plot / tradeoff_plot) +
  patchwork::plot_annotation(
    title = "Figure 1. Actual-mock benchmarking reveals method- and threshold-dependent performance",
    tag_levels = "A"
  )

print(figure_1)
ggsave(
  file.path(output_dir, "07_Figure1_Karstens_mock_benchmark.png"),
  figure_1, width = 16, height = 12, dpi = 300
)
ggsave(
  file.path(output_dir, "07_Figure1_Karstens_mock_benchmark.pdf"),
  figure_1, width = 16, height = 12
)
```

# 7. Method-by-intensity comparison

```{r comparison-heatmap, fig.width=11, fig.height=6}
comparison_plot_data <- focus_summary |>
  mutate(
    label = sprintf("%.2f", mean_Youden_J),
    method = factor(method, levels = rev(names(method_colors)))
  )

p_comparison <- ggplot(
  comparison_plot_data,
  aes(intensity, method, fill = mean_Youden_J)
) +
  geom_tile(color = "white", linewidth = 0.7) +
  geom_text(aes(label = label), size = 4) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    midpoint = 0, limits = c(-1, 1)
  ) +
  labs(
    x = NULL, y = NULL, fill = "Mean Youden J",
    title = "No single removal intensity is optimal for every method",
    subtitle = "Mean over the published focus dilutions D2, D4, D6, and D8"
  )

print(p_comparison)
ggsave(
  file.path(output_dir, "07_method_intensity_Youden_heatmap.png"),
  p_comparison, width = 11, height = 6, dpi = 300
)
```

# 8. Verification and interpretation boundary

```{r verification}
verification <- data.frame(
  check = c(
    "Official dataset contains 10 samples and 1414 ASVs",
    "Gold standard contains 9 expected and 1405 contaminant ASVs",
    "Five methods and three intensity levels completed",
    "Every method-intensity combination produced all nine dilution results",
    "Confusion matrices recover every original sample read",
    "All reported metrics are within their valid ranges"
  ),
  passed = c(
    nsamples(ps) == 10L & ntaxa(ps) == 1414L,
    sum(!truth_table$is_contaminant) == 9L & sum(truth_table$is_contaminant) == 1405L,
    n_distinct(benchmark_metrics$method) == 5L &
      n_distinct(benchmark_metrics$intensity) == 3L,
    nrow(benchmark_metrics) == 5L * 3L * 9L,
    all(
      benchmark_metrics$TP + benchmark_metrics$TN +
        benchmark_metrics$FP + benchmark_metrics$FN ==
        rowSums(count_matrix[as.character(benchmark_metrics$SampleID), , drop = FALSE])
    ),
    all(benchmark_metrics$Sensitivity_Recall >= 0 & benchmark_metrics$Sensitivity_Recall <= 1) &
      all(benchmark_metrics$Specificity >= 0 & benchmark_metrics$Specificity <= 1) &
      all(benchmark_metrics$Precision_PPV >= 0 & benchmark_metrics$Precision_PPV <= 1) &
      all(benchmark_metrics$F1 >= 0 & benchmark_metrics$F1 <= 1) &
      all(benchmark_metrics$Youden_J >= -1 & benchmark_metrics$Youden_J <= 1)
  )
)

write_csv(verification, file.path(output_dir, "07_06_verification.csv"))
knitr::kable(verification)
if (!all(verification$passed)) stop("One or more benchmark checks failed.")
```

This even mock is a controlled benchmark, not a universal threshold selector.
It tests preservation of nine known ASVs under a biomass gradient. A threshold
that performs well here may still remove genuine inhabitants in a real skin or
other low-biomass dataset. The next validation layer should therefore combine
negative-control enrichment with explicit retention of literature-supported
commensal genera before a threshold is accepted for biological analysis.

```{r session-info}
sessionInfo()
```
